---
title: "Stat 343 PSet 6"
author: "Karl Jiang"
date: "November 17, 2017"
output: pdf_document
---

```{r, include = FALSE}
library(alr4)
library(faraway)
```


#1) MBA Students 

##a) 
```{r}
data("happy")
mdl_happy = lm(happy ~ ., data = happy)
summary(mdl_happy)
```

Without correcting for the FWER, using $\alpha = 0.01$ the only significant predictor is "love". 

##b) "Questionable"


```{r}
table(happy$happy)
```

To perform t tests, we assume that the errors are normally distributed (continous values on support $[-\infty, \infty]$ with constant variance $\sigma^2$. Because the response is a discrete, there's no way the errors can be normally distributed. 

##c) Permutation Money 


```{r, cache = TRUE}
n_iter = 1000
money_coeffs = c() 
for(i in 1:n_iter){
  mdl_tmp = lm(happy ~ . -money + sample(money), data = happy)
  s = summary(mdl_tmp)
  money_coeffs = c(money_coeffs, s$coefficients["sample(money)", "t value"] )
}

hist(money_coeffs, probability = TRUE, main = "Money T Stat Permutations")

#Compute the perumtated p value
t_money = summary(mdl_happy)$coeff["money", "t value"]
sprintf("the two sided p value is: %f", mean(abs(money_coeffs) > t_money))
```

##d) Density over Histogram 

```{r}
grid = seq(-4, 4, length = 3000)
t_vals = dt(grid, df = mdl_happy$df.residual)
hist(money_coeffs, probability = TRUE, main = "Money T Stat Permutations")
lines(t_vals ~ grid, type = "l")
```


##e) Bootstrap 

```{r, cache = TRUE}
res = mdl_happy$residuals
y_est = mdl_happy$fitted.values

b_money = summary(mdl_happy)$coeff["money", "Estimate"]

B = 5000
boot_money = c()
for(i in 1:B){
  e_star = res[sample(length(res), replace = T)]
  ynew = y_est + e_star
  mdl_tmp_s = summary(lm(ynew ~ ., data = happy))
  b_money = c(b_money, mdl_tmp_s$coefficients["money", "Estimate"])
}
```

```{r}
#CI 
#hist(b_money)
print("95 Boostrap CI: " )
print( quantile(b_money,c(0.025,0.975)) ) 

print("90% Bootstrap CI: ")
print( quantile(b_money,c(0.05,0.95)) ) 
```

As we can see, 0 does not fall in the $1 - \alpha = 95\%$ confidence interval. However, 0 does fall within the 90% confidence interval. This is consistent with the p value from the vanilla MLR, which is significant at the $\alpha = 0.1$ level (eg. 0 not in 90% CI), but not at the $\alpha = 0.05$ level. 

#2) ToothGrowth, Factor levels

(Text book, Chapter 14, Problem 3) Plot the ToothGrowth data with len as
the response. Fit a linear model to the data and check for possible simplification.
Display the fitted regression lines on top of the data.


Since the predictor supp is binary, we shall display two plots for each category (OJ and VC) along the values of dose. Because there are only 3 unique values of dosage levels, we should treat this as a categorical. However, looking at the fit below (using interaction term between dosage and supp), a possible simplication - treating dosage as continuous, could be possible here, as each unit increase in dosage seems to correspond to a linear change in length. 

```{r}
data("ToothGrowth")
#summary(ToothGrowth)

#lm
mdl_t = lm(len ~ . + supp * dose, data = ToothGrowth)
summary(mdl_t)

tj_oj = ToothGrowth[ToothGrowth$supp == "OJ", ]
tj_vc = ToothGrowth[ToothGrowth$supp == "VC", ]

fit_oj = predict(mdl_t, newdata = tj_oj)
fit_vc = predict(mdl_t, newdata = tj_vc)
#seperate 
plot(len ~ dose, data = ToothGrowth, subset = supp == "OJ", col = "red", 
     main = "OJ")
lines(fit_oj ~ tj_oj$dose )

plot(len ~ dose, data = ToothGrowth, subset = supp == "VC", col = "red", 
     main = "VC")
lines(fit_vc ~ tj_vc$dose )
```

```{r}
#together 
plot(len ~ dose, data = ToothGrowth, subset = supp == "OJ", col = "red", 
     main = "Red = OJ, Blue = VC")
points(len ~ dose, data = ToothGrowth, subset = supp == "VC", pch = 5, col = "blue")
lines(fit_oj ~ tj_oj$dose )
lines(fit_vc ~ tj_vc$dose )
```

Both fits (on OJ and VC) treating dosage as a continuos variable seem to fit the data quite well (Solid $R^2$ and significant predictor in dosage treated as a continuos variable). 

Moreover, to further test if we can do this simplication, we can do a hypothesis test. With $\beta_{dosage = 0.5}$ as the default, our null is $H_0: \beta_{dosage = 1} = 2\beta_{dosage = 2}$, which suggests that the change is linear in dosage (at least to 1.0 and 2.0, not actually a conclusive test). 